---
title: "Analysis of IIJA Funding Allocation"
author: "Souleymane Doumbia"
date: "`r Sys.Date()`"
output:
  slidy_presentation: 
    fig_width: 7
    fig_height: 5
    keep_md: true
    toc: true
    toc_depth: 3
    toc_float: true
  beamer_presentation:
    fig_width: 7
    fig_height: 5
    keep_md: true
    toc: true
  ioslides_presentation:
    fig_width: 7
    fig_height: 5
    keep_md: true
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Required Libraries
library(ggplot2)
library(dplyr)
library(reshape2)
library(ggthemes)
library(corrplot)
```

# Introduction
This report examines the allocation of the Infrastructure Investment and Jobs Act (IIJA) funding across U.S. states and territories. The analysis addresses two key questions:

1. Is the allocation equitable based on the population and infrastructure needs of the states, or is there evidence of bias?

2. Does the allocation favor the political interests of the Biden administration?


# Data Overview
The dataset includes:

- IIJA funding allocation per state as of March 2023

- Population estimates as of July 1, 2023

- Infrastructure grades from the ASCE

- Unemployment rates as of March 2023

- Political affiliation of each state based on the 2020 presidential election


```{r Data_Preparation, echo=FALSE}
# Data Preparation & Analysis:
# Loading the dataset
data <- read.csv("https://raw.githubusercontent.com/Doumgit/D608-Projects/refs/heads/main/IIJA%20FUNDING%2003_2023%20.csv")  

# Clean up the population column (convert to numeric by removing commas)
data$Population.Estimate.as.July.1..2023.USCENSUS <- as.numeric(gsub(",", "", data$Population.Estimate.as.July.1..2023.USCENSUS))

# Converting other necessary columns to numeric
data$Total..Billions. <- as.numeric(data$Total..Billions.)
data$Unemployment.rate.03.2023 <- as.numeric(data$Unemployment.rate.03.2023)

# Filtering out rows with missing values 
data_clean <- data %>%
  filter(!is.na(Total..Billions.) & !is.na(Population.Estimate.as.July.1..2023.USCENSUS) & State..Teritory.or.Tribal.Nation != 'PUERTO RICO')

# Calculating Per Capita Funding
data_clean$Per_Capita_Funding <- data_clean$Total..Billions. * 1e9 / data_clean$Population.Estimate.as.July.1..2023.USCENSUS

# Grouping unemployment rates into categories
data_clean$Unemployment_Group <- cut(data_clean$Unemployment.rate.03.2023, breaks = c(0, 2, 4, 6, 8), labels = c("0-2%", "2-4%", "4-6%", "6-8%"))

# Summarizing funding by political affiliation
party_funding <- data_clean %>% 
  group_by(X2020.Presidential.Winner) %>%
  summarise(Total_Funding = sum(Total..Billions.))

# Prepare correlation matrix
corr_data <- data_clean %>%
  select(Per_Capita_Funding, Population.Estimate.as.July.1..2023.USCENSUS, Unemployment.rate.03.2023)
corr_matrix <- cor(corr_data, use = "complete.obs")
```


# Proportion of Total Funding by Political Affiliation:
```{r Proportion_of_Total_Funding_by_Political_Affiliation, echo=FALSE}
ggplot(party_funding, aes(x = "", y = Total_Funding, fill = X2020.Presidential.Winner)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  theme_void() +
  labs(title = "Proportion of Total Funding by Political Affiliation", fill = NULL) +
  scale_fill_manual(values = c("blue", "red")) +
  theme(legend.position = "bottom")
```



# Total Funding by State and Infrastructure Grade:
```{r Total_Funding_by_State_and_Infrastructure_Grade, echo=FALSE}
## 2. Total Funding by State and Infrastructure Grade
ggplot(data_clean, aes(x = reorder(State..Teritory.or.Tribal.Nation, -Total..Billions.), y = Total..Billions., fill = X2021.ASCE.Infrastructure.Grading.by.State)) +
  geom_bar(stat = "identity", width = 1) +
  coord_flip() +  # Flip the axes
  labs(title = "Total Funding by State and 2021 ASCE Infrastructure Grade",
       x = "State",
       y = "Total Funding (Billions)",
       fill = "2021 ASCE Infrastructure Grade") +  # Corrected legend title
  theme_minimal() +  # Use minimal theme
  theme(axis.text.y = element_text(angle = 0, hjust = 1, size = 6),  # Make x-axis labels vertical
        axis.text.x = element_text(size = 8),  # Increase label size for state names
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10),
        legend.title = element_text(size = 10),
        legend.position = "right",
        panel.grid = element_blank()) +  # Remove grid lines
  scale_fill_brewer(palette = "Set3")
```



# Per Capita Funding by Political Affiliation:
```{r Per_Capita_Funding_by_Political_Affiliation, echo=FALSE}
## 3. Per Capita Funding by Political Affiliation
ggplot(data_clean, aes(x = X2020.Presidential.Winner, y = Per_Capita_Funding)) +
  geom_boxplot() +
  labs(title = "Per Capita Funding by Political Affiliation",
       x = "Political Affiliation",
       y = "Per Capita Funding (USD)") +
  theme_minimal() +  # Keep the minimal theme
  theme(plot.title = element_text(hjust = 0.5),
        panel.grid = element_blank(),  # Remove all grid lines
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10),
        axis.text = element_text(size = 9))  # Adjust label sizes if needed
```



# Correlation Matrix:
```{r Correlation_Matrix, echo=FALSE}
colnames(corr_matrix) <- c("Per Capita", "Population", "Unemployment")
rownames(corr_matrix) <- c("Per Capita", "Population", "Unemployment")

#plotting correlation matrix
corrplot(corr_matrix, method = "circle", type = "upper",
         title = "Correlation Matrix of Key Variables",
         tl.col = "black", tl.srt = 45,  
         tl.cex = 1, 
         mar = c(0, 0, 2, 0)
         )
```



# Conclusion
The analysis reveals:

- **No clear partisan bias:** Republican states receive more per capita funding despite being fewer in number, which indicates no favoritism toward Democratic states.

- **Weak correlation between funding and infrastructure or unemployment:** Infrastructure needs and unemployment rates are not strongly correlated with funding, suggesting other factors may influence the allocation.

- **Funding allocation based on per capita differences:** States vary significantly in their per capita funding, with Republican states tending to receive more.


# Sources:
- **U.S. BUREAU OF LABOR STATISTICS (BLS)** for Unemployment Rates per State**:** https://www.bls.gov/lau/

- **AMERICAN SOCIETY OF CIVIL ENGINEERS (ASCE)** for Infrastructure Grades per State**:** https://infrastructurereportcard.org

- **UNITED STATES CENSUS BUREAU** for Population Estimates per State**:** https://www.census.gov/data/tables/time-series/demo/popest/2020s-state-total.html

- **THE NEW YORK TIMES** for 2020 Presidential Election Results**:** https://www.nytimes.com/interactive/2020/11/03/us/elections/results-president.html?action=click&pgtype=Article&state=default&module=styln-elections-2020&region=TOP_BANNER&context=election_recirc

