---
title: "DATA 608 Story 2"
author: "Souleymane Doumbia"
date: "`r Sys.Date()`"
output:
  beamer_presentation:
    fig_width: 7
    fig_height: 5
    keep_md: true
    toc: true
  slidy_presentation:
    fig_width: 7
    fig_height: 5
    keep_md: true
    toc: true
    toc_depth: 3
    toc_float: true
  ioslides_presentation:
    fig_width: 7
    fig_height: 5
    keep_md: true
    toc: true
    toc_depth: 3
  html_document:
    toc: true
    toc_depth: '3'
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Required Libraries
library(ggplot2)
library(dplyr)
library(lubridate)
library(yaml)
library(httr)
library(jsonlite)
library(caret)
library(fredr)
library(blscrapeR)
```

## Introduction

The Federal Reserve's mandate, given by Congress, focuses on two primary goals:

- **Control inflation**: Measured using the Consumer Price Index (CPI), which tracks price changes over time.
- **Maintain low unemployment**: Tracked through the unemployment rate.

At times, these goals can appear conflicting. For instance, raising interest rates to control inflation may result in slower economic growth, which can increase unemployment. This project explores data from the last 25 years to examine how the Federal Reserve has attempted to balance these two objectives.

## Introduction
We used the following datasets:

- **Consumer Price Index (CPI)** - Sourced from the Bureau of Labor Statistics (BLS) using BLS API
- **Unemployment Rate** - Also sourced from BLS using BLS API
- **Federal Funds Rate** - Obtained from the Federal Reserve's FRED database using Fred API

**Key Questions:**

1. **Has the Fed effectively controlled inflation over the last two decades?**
2. **How has the unemployment rate been impacted by changes in the Fed Funds Rate?**
3. **Is there evidence of a relationship between inflation and unemployment (Phillips Curve)?**

This Report includes multiple Visualizations and statistical insights to help answer these questions.



```{r  1_The_FED Funds_Rate_FRED_Federal_Reserve_Board, echo=FALSE}
library(fredr)
library(yaml)


# Reading the Fred API key from the YAML file
config <- yaml::read_yaml("fred_api.yml")
fred_api_key <- config$FRED_API_KEY

# Setting FRED API key
fredr_set_key(fred_api_key)

# Retrieving the Federal Funds Rate data
fed_funds_rate <- fredr(
  series_id = "FEDFUNDS",
  observation_start = as.Date("1999-01-01"),
  observation_end = as.Date("2024-12-31")
)

# Viewing the data
#head(fed_funds_rate)
```


```{r 2_The_Consumer_Price_Index_CPI_Bureau_of_Labor_Statistics, echo=FALSE}
library(blscrapeR)

# Reading the BLS API key from the YAML file
config <- yaml::read_yaml("bls_api.yml")
bls_api_key <- config$BLS_API_KEY

# Fetching Cinsumer Price Index (CPI) data from 1999 to 2010
cpi_1999_2010 <- bls_api(
  "CUSR0000SA0", 
  startyear = 1999, 
  endyear = 2010,
  registrationKey = bls_api_key
)

# Fetching Consumer Price Index (CPI) data from 2011 to 2024
cpi_2011_2024 <- bls_api(
  "CUSR0000SA0", 
  startyear = 2011, 
  endyear = 2024,
  registrationKey = bls_api_key
)
cpi_2011_2024 <- cpi_2011_2024 %>%
  select(year, period, periodName, value, footnotes, seriesID)

# Combining both periods
consumer_price_index <- rbind(cpi_1999_2010, cpi_2011_2024)

# Removing any duplicates
#consumer_price_index <- consumer_price_index[!duplicated(consumer_price_index$year), ]


# Viewing the first few rows
#head(consumer_price_index)
```


```{r 3_Unemployment_Rate_Bureau_of_Labor_Statistics, echo=FALSE}
# Fetching unemployment rate data using the correct API key parameter
unemployment_1999_2010 <- bls_api(
  "LNS14000000",  # Series ID for Unemployment Rate
  startyear = 1999, 
  endyear = 2010, 
  registrationKey = bls_api_key  # Using the API key with the 'registrationKey' parameter
)

unemployment_2011_2024 <- bls_api(
  "LNS14000000",  # Series ID for Unemployment Rate
  startyear = 2011, 
  endyear = 2024, 
  registrationKey = bls_api_key  # Using the API key with the 'registrationKey' parameter
)
unemployment_2011_2024 <- unemployment_2011_2024 %>% select(-latest)

unemployment_rate <- rbind(unemployment_1999_2010, unemployment_2011_2024)

# Viewing the first few rows of unemployment data
#head(unemployment_rate)
```

```{r 4_Data_Preparation_Merging_all_three_datasets, echo=FALSE}
library(dplyr)
library(lubridate)

# 1. Converting CPI data to a proper date format
consumer_price_index <- consumer_price_index %>%
  mutate(date = as.Date(paste0(year, "-", substr(period, 2, 3), "-01")))  # Creating date column

# 2. Converting Unemployment data to a proper date format
unemployment_rate <- unemployment_rate %>%
  mutate(date = as.Date(paste0(year, "-", substr(period, 2, 3), "-01")))  # Creating date column

# 3. Selecting relevant columns from each dataset
fed_funds_rate_clean <- fed_funds_rate %>%
  select(date, Fed_Funds_Rate = value)  # Fed Funds Rate

cpi_clean <- consumer_price_index %>%
  select(date, CPI = value)  # CPI

unemployment_clean <- unemployment_rate %>%
  select(date, Unemployment_Rate = value)  # Unemployment Rate

# 4. Merging all datasets by the date column
merged_data <- fed_funds_rate_clean %>%
  left_join(cpi_clean, by = "date") %>%
  left_join(unemployment_clean, by = "date")

# View the merged data
#head(merged_data)
```


```{r 5_Normalizing_the_Data, echo=FALSE}
# Normalizeing CPI, Fed Funds Rate, and Unemployment Rate
merged_data <- merged_data %>%
  mutate(CPI_scaled = (CPI - min(CPI, na.rm = TRUE)) / (max(CPI, na.rm = TRUE) - min(CPI, na.rm = TRUE)),
         Fed_Funds_Rate_scaled = (Fed_Funds_Rate - min(Fed_Funds_Rate, na.rm = TRUE)) / (max(Fed_Funds_Rate, na.rm = TRUE) - min(Fed_Funds_Rate, na.rm = TRUE)),
         Unemployment_Rate_scaled = (Unemployment_Rate - min(Unemployment_Rate, na.rm = TRUE)) / (max(Unemployment_Rate, na.rm = TRUE) - min(Unemployment_Rate, na.rm = TRUE)))

# View the normalized data
#head(merged_data)
```


## 1. Relationships between the Fed Funds Rate, CPI, and Unemployment Rate
```{r Relationships_between_the_Fed_Funds_Rate_CPI_and_Unemployment_Rate, echo=FALSE}
library(ggplot2)

suppressWarnings({
  # Plotting the normalized CPI, Fed Funds Rate, and Unemployment Rate
ggplot(merged_data, aes(x = date)) +
  geom_line(aes(y = CPI_scaled, color = "CPI"), size = 1) +
  geom_line(aes(y = Unemployment_Rate_scaled, color = "Unemployment Rate"), size = 1) +
  geom_line(aes(y = Fed_Funds_Rate_scaled, color = "Fed Funds Rate"), size = 1) +
  labs(title = "Normalized CPI, Unemployment Rate, and Fed Funds Rate (1999-2024)",
       x = "Year", y = "Normalized Value") +
  scale_color_manual(values = c("CPI" = "red", "Unemployment Rate" = "blue", "Fed Funds Rate" = "green")) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    legend.position = "bottom",
    legend.title = element_blank(),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),  # Centered title
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )
})


```


## 2. Key Economic Events
```{r Key_Economic_Events, echo=FALSE}
# Plotting the normalized CPI, Fed Funds Rate, and Unemployment Rate with annotations and shaded regions
ggplot(merged_data, aes(x = date)) +
  # Plottinh the lines for the indicators
  geom_line(aes(y = CPI_scaled, color = "CPI"), size = 1) +
  geom_line(aes(y = Unemployment_Rate_scaled, color = "Unemployment Rate"), size = 1) +
  geom_line(aes(y = Fed_Funds_Rate_scaled, color = "Fed Funds Rate"), size = 1) +
  
  # Highlighting shaded regions for economic events
  geom_rect(aes(xmin = as.Date("2000-03-01"), xmax = as.Date("2002-12-01"), ymin = -Inf, ymax = Inf), fill = "gray", alpha = 0.006) +  # Dot-com Bubble
  geom_rect(aes(xmin = as.Date("2007-12-01"), xmax = as.Date("2009-06-01"), ymin = -Inf, ymax = Inf), fill = "gray", alpha = 0.006) +  # Financial Crisis
  geom_rect(aes(xmin = as.Date("2020-02-01"), xmax = as.Date("2021-06-01"), ymin = -Inf, ymax = Inf), fill = "gray", alpha = 0.006) +  # COVID-19 Pandemic
  geom_rect(aes(xmin = as.Date("2022-01-01"), xmax = as.Date("2024-12-31"), ymin = -Inf, ymax = Inf), fill = "gray", alpha = 0.006) +  # Inflation Surge
  
  # Adding annotations for significant economic events
  annotate("text", x = as.Date("2001-06-01"), y = 0.7, label = "Dot-com Bubble", color = "black", angle = 90, vjust = -0.5) +
  annotate("text", x = as.Date("2009-06-01"), y = 0.7, label = "Financial Crisis", color = "black", angle = 90, vjust = -0.5) +
  annotate("text", x = as.Date("2020-06-01"), y = 0.7, label = "COVID-19 Pandemic", color = "black", angle = 90, vjust = -0.5) +
  annotate("text", x = as.Date("2023-01-01"), y = 0.7, label = "Inflation Surge", color = "black", angle = 90, vjust = -0.5) +
  
  # Setting axis labels and titles
  labs(title = "Normalized CPI, Unemployment Rate, and Fed Funds Rate (1999-2024)",
       x = "Year", y = "Normalized Value") +
  scale_color_manual(values = c("CPI" = "red", "Unemployment Rate" = "blue", "Fed Funds Rate" = "green")) +
  
  # Minimal theme settings
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    legend.position = "bottom",
    legend.title = element_blank(),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),  # Centering title
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )

```

## 3. Scatter Plot of Inflation vs. Unemployment (Phillips Curve)
```{r Scatter_Plot_of_Inflation_vs_Unemployment_Phillips_Curve_Analysis, message=FALSE, echo=FALSE}
# scatter plot of CPI vs. Unemployment Rate (Phillips Curve analysis)
suppressWarnings(
  ggplot(merged_data, aes(x = CPI, y = Unemployment_Rate)) +
    geom_point(color = "blue", alpha = 0.6) +
    geom_smooth(method = "lm", color = "red", se = FALSE, show.legend = FALSE) +  # Suppressing formula message
    labs(title = "Relationship Between Inflation (CPI) and Unemployment Rate",
         x = "CPI (Inflation)", y = "Unemployment Rate") +
    theme_minimal() +
    theme(
      panel.grid = element_blank(),  # Removing grid lines
      plot.title = element_text(size = 14, hjust = .6),
      axis.title = element_text(size = 12),
      axis.text = element_text(size = 10)
    ) +
    annotate("text", x = 270, y = 12, label = "High Unemployment \nPeriod (2009)", color = "red", vjust = -0.5, fontface = "bold")  # Example annotation
)
```


## 4. Bar Plot of Average Fed Funds Rate, CPI, and Unemployment During Key Economic Events
```{r Bar_Plot_of_Average_Fed_Funds_Rate_CPI_and_Unemployment_During_Key_Economic_Events, echo=FALSE}
# Creating a summary dataset to calculate the average values during key events
event_summary <- merged_data %>%
  mutate(event = case_when(
    date >= as.Date("2000-03-01") & date <= as.Date("2002-12-01") ~ "Dot-com Bubble",
    date >= as.Date("2007-12-01") & date <= as.Date("2009-06-01") ~ "Financial Crisis",
    date >= as.Date("2020-02-01") & date <= as.Date("2021-06-01") ~ "COVID-19 Pandemic",
    date >= as.Date("2022-01-01") ~ "Inflation Surge",
    TRUE ~ NA_character_
  )) %>%
  group_by(event) %>%
  summarise(avg_CPI = mean(CPI, na.rm = TRUE),
            avg_Unemployment = mean(Unemployment_Rate, na.rm = TRUE),
            avg_Fed_Funds_Rate = mean(Fed_Funds_Rate, na.rm = TRUE)) %>%
  filter(!is.na(event))

# Bar plot with annotations
ggplot(event_summary, aes(x = event)) +
  geom_bar(aes(y = avg_CPI, fill = "CPI"), stat = "identity", position = "dodge") +
  geom_bar(aes(y = avg_Unemployment * 10, fill = "Unemployment Rate"), stat = "identity", position = "dodge") +
  geom_bar(aes(y = avg_Fed_Funds_Rate * 10, fill = "Fed Funds Rate"), stat = "identity", position = "dodge") +
  labs(title = "Average Fed Funds Rate, CPI, and Unemployment Rate During Key Economic Events",
       x = "Economic Event", y = "Average Value (Scaled)") +
  scale_fill_manual(values = c("CPI" = "red", "Unemployment Rate" = "blue", "Fed Funds Rate" = "green")) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),  # Removing grid lines
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    legend.position = "bottom"
  ) +
  guides(fill = guide_legend(title = NULL)) +  # Removing "fill" from the legend

  # Adding annotations
  annotate("text", x = 1, y = 140, label = "High Unemployment Spike", color = "black", size = 3, angle = 90, hjust = -0.2) +  
  annotate("text", x = 2, y = 70, label = "Low Fed Rate to Stimulate Economy", color = "black", size = 3, angle = 90, hjust = -0.2) +  
  annotate("text", x = 3, y = 140, label = "Severe Unemployment", color = "black", size = 3, angle = 90, hjust = -0.2) +  
  annotate("text", x = 4.1, y = 120, label = "Higher Rates to Curb Inflation", color = "black", size = 3, angle = 90, hjust = -0.2) +
  annotate("text", x = 4, y = 120, label = "High CPI, Rising Inflation", color = "black", size = 3, angle = 90, hjust = -0.2)
```

## 5. Correlation Heatmap Between CPI, Unemployment Rate, and Fed Funds Rate
```{r Correlation_Heatmap_Between_CPI_Unemployment_Rate_and_Fed_Funds_Rate, echo=FALSE}
# Correlation Matrix
cor_matrix <- cor(merged_data[, c("CPI", "Unemployment_Rate", "Fed_Funds_Rate")])

# Convertin correlation matrix into a long format for ggplot
cor_data <- as.data.frame(as.table(cor_matrix))

# Plotting the heatmap with correlation coefficients inside the squares
ggplot(cor_data, aes(Var1, Var2, fill = Freq)) +
  geom_tile(color = "white") + 
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, limit = c(-1, 1), space = "Lab", name = "Correlation") +
  geom_text(aes(label = round(Freq, 2)), size = 5, color = "black") +  # Adding text labels for correlation coefficients
  labs(title = "Correlation Heatmap: CPI, Unemployment Rate, and Fed Funds Rate",
       x = "", y = "") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank(),  # Removing background grid
        plot.title = element_text(size = 14, face = "bold", hjust = 0.5))
```


## Summary
- **Visual 1 & 2 - Trend Analysis (Normalized CPI, Unemployment Rate, and Fed Funds Rate)**: 
  - These charts demonstrate how the Fed has raised and lowered interest rates in response to economic events. 
  - For example, during the **COVID-19 Pandemic**, we see a sharp increase in unemployment, followed by a low **Fed Funds Rate** to stimulate economic recovery.
  - As inflation surged in **2021-2022**, the Fed responded by raising rates dramatically to curb rising prices.

- **Visual 2 - Relationship Between Inflation and Unemployment (Phillips Curve)**:
  - The scatter plot shows a weak negative correlation between **inflation** and **unemployment**, as expected based on the **Phillips Curve**.
  - However, exceptions occur during extreme economic crises, such as in **2009 (Great Recession)** when unemployment spiked despite relatively stable inflation.
  - This suggests that the traditional relationship between inflation and unemployment may weaken or invert during economic shocks.

## Summary
- **Visual 3 - Bar Plot of Average CPI, Unemployment, and Fed Funds Rate by Event**:
  - During events like the **COVID-19 Pandemic** and the **Financial Crisis**, **unemployment** was high, with low **Fed Funds Rates**, while **inflation (CPI)** remained relatively stable.
  - Conversely, during the **Inflation Surge (2021-2022)**, the Fed aggressively raised interest rates to combat rising inflation, and we see **CPI** towering above other metrics during this period.

- **Visual 4 - Correlation Heatmap**:
  - The correlation heatmap reveals that the **Fed Funds Rate** and **Unemployment Rate** are negatively correlated (-0.6), while the correlation between **CPI** and both the **Fed Funds Rate** (-0.17) and **Unemployment Rate** (-0.14) is weakly negative.
  - This suggests that as the **Fed Funds Rate** increases, unemployment tends to decrease, as higher rates often coincide with efforts to curb inflation and maintain economic stability.
  - Conversely, when the **Fed Funds Rate** decreases (often during economic crises), unemployment tends to rise, as seen in periods of recovery when interest rates are kept low to stimulate growth, but unemployment remains high until full recovery.


## Conclusion

- The **Fed** has effectively utilized the **Fed Funds Rate** to control inflation during periods of rising prices (e.g., 2021-2022).
- The relationship between **inflation** and **unemployment** appears weaker during economic crises (e.g., 2008, 2020).
- The Fed's dual mandate remains a challenging balancing act, as actions to reduce inflation can adversely affect unemployment, particularly during crises where both **inflation** and **unemployment** can rise simultaneously.
- The analysis demonstrates that while the **Federal Reserve** has been largely effective in responding to economic crises with appropriate rate adjustments, the interaction between inflation and unemployment is complex, and external shocks often require adaptive and non-linear responses.

## Sources
- **Consumer Price Index (CPI)** - Bureau of Labor Statistics (BLS): [https://www.bls.gov](https://www.bls.gov)
- **Unemployment Rate** - Bureau of Labor Statistics (BLS): [https://www.bls.gov](https://www.bls.gov)
- **Federal Funds Rate** - Federal Reserve's FRED Database: [https://fred.stlouisfed.org](https://fred.stlouisfed.org)
